\racets: Faceted Execution in Racket

### 2018-07-24 paper/paper.tex
adding fac cite
self-cite

To date, implementations of faceted execution are relatively
heavyweight: requiring either changing the language runtime or the
application code (e.g., by using monads). Following Racket's
languages-as-libraries approach, we present Racets: an implementation
of faceted execution as a library of macros.  Given Racket's
highly-expressive macro system, our implementation follows relatively
directly from the semantics of faceted execution. To demonstrate how
\racets can be used for policy-agnostic programming, we use it to
build a web-based game of Battleship. Our implementation sheds light
on several interesting issues in interacting with code written without
faceted execution. Our \racets implementation is open source, under
development, and available online.
\end{abstract}

@@

Policy-agnostic programming promotes the idea that programmers should
be able to write programs ``normally'', without concerning themselves
with how privacy policies are enforced. Unfortunately, the
relatively foundational linguistic changes required to enable faceted
execution have hindered implementations thusfar. Dynamic generation
of first-class security labels, tracking an arbitrary number of
facets per value, and keeping faceted-value trees in a canonical order,
are all central challenges in any practical implementation. For example, the
first implementation of FE (by \citet{Austin:2012}) extended
a JavaScript interpreter to account for faceted values. Other
implementations use monads~\cite{Schmitz:2016}
or rely upon third-party macro systems~\cite{Yang:2016}. We
know of no existing implementation of FE that aims to interoperate seamlessly
with code written in the host language. By contrast, Scheme boasts a powerful
hygienic macro system that allows essentially any linguistic form
to be modified arbitrarily.

In this paper we present \racets, an implementation of policy-agnostic
programming in Racket via macros. \racets provides facilities for
creating policies and faceting secure data with those
policies. \racets also extends several core forms in Racket to work
with faceted values (our implementation is detailed in
Section~\ref{sec:implementation}). We have used \racets to implement a
small server-based board-game (detailed in Sections~\ref{sec:overview}
and~\ref{sec:implementation}). Relevant related work is presented in
section~\ref{sec:related}. We see \racets as a promising prototype for
policy-agnostic programming in Racket, and conclude with discussion of
future directions in Section~\ref{sec:conclusion}.

@@

To introduce faceted execution more concretely, we present the implementation
of Battleship, a small guessing game, in \racets (this section presents a
distilled version of our case study in
Section~\ref{sec:implementation}). In this game each player has a
private board of coordinates, at which they place ``ships''. The players
hide their boards from each other as play progresses in rounds. Each
turn a player guesses the position of a ship on the other player's
board. If the guess is successful the tile is removed from the
board and a hit is declared publicly. Play ends once one player's board
has no remaining tiles, at which point that player loses.

@@

In the above example, \blue{$\star$} (lazy failure) is used in the negative branch
to ensure execution will fail if Bob tries to observe Alice's secret gameboard.
The $\code{(obs}\ \blue{e_\ell}\ \blue{e_\text{key}}\ \blue{e_\text{fac}}\code{)}$ form takes
a label, a key, and a faceted value. Explicit observation projects a single label
$\blue{e_\ell}$ in faceted value $\blue{e_\text{fac}}$ to either its positive or negative facet,
depending on whether the policy associated with $\blue{e_\ell}$ returns true for key
$\blue{e_\text{key}}$. If Bob tries to observe Alice's board, the policy predicate
will return false (from \texttt{(= 1 $\blue{2}$)}) and Alice's negative facet $\blue{\star}$ will result.

@@

Because the applied function can be stateful, faceted execution also
maintains the current privilege level in a \emph{program counter} (PC). The
program counter is a property of the current evaluation context and is used
to build facets when writes are made to the store in a privileged context.
For example, if a stateful function ``splits'' when applied on both the positive and negative
facets of a value faceted by a label $\blue{\ell}$, and on the positive branch
the function uses $\texttt{set!}$ to mutate a variable $x$ from $2$ to $3$,
FE semantics will set $x$ to $\facet{\ell}{3}{2}$ so that the value $3$
cannot be leaked from the secure context (speculative execution under the
$\blue{+\ell}$ facet). This is because, for the duration of the the app ``split'',
the evaluation context records that all values are implicitly guarded by $\blue{+\ell}$
and then $\blue{-\ell}$, respectively. If the semantics for $\texttt{set!}$ does
not make this faceting explicit, a sensitive value could leak from one PC to another.
We expand upon these subtleties in Section~\ref{sec:semantics},
where we present a complete semantics for faceted execution.

@@

Our semantics is shown in Figure~\ref{fig:semantics}. As \lang is an
extension of the lambda calculus with references, we present the parts
unique to faceted execution in \fcol{red}, while keeping the lambda
calculus with references in \blue{blue}. Base values in our semantics
include addresses (used for boxes), constants, and closures. We also
include a kind of lazy failure (\bmth{\star}), which is necessary for
defining store update within a protected context.

@@

The rules \textsc{Box}, \textsc{Unbox}, and \textsc{Set} appear
similar to the standard implementation of boxes, but employ several
meta-functions to do their work. This is because box creation, reads,
and writes may occur within a privileged context, and care must be
taken to form facets when \rmth{pc} is nonempty. To understand why,
consider the following example\footnote{Our formal semantics elides
  \code{if}, though it may be obtained via a Church encoding if
  desired as in Austin et al.\cite{Austin:2012}. Our implementation
  includes \code{if} but not other constructs such as \code{cond}}:

@@

Facet creation appears as three separate rules in
Figure~\ref{fig:semantics}, which we recapitulate here in three
distinct colors for each case:

@@

Our hypothesis is that this tagging scheme is key to permitting inter-operation
between \racets and Racket, and we have implemented a faceted, web-based game of Battleship
to explore this idea. We suspect that a more thorough investigation of likely idioms
for faceted, non-faceted module interaction is needed and that purely functional code
plays a special role as a degenerate case where arbitrary non-faceted code may be lifted
to operate over faceted values without potential unsoundness. In the future, we plan to explore
these design choices in a more principled way and also to scale a static analysis of faceted
execution to fully expanded Racket so it may be applied directly to \racets.

>>>>

To date, implementations of faceted execution are relatively
heavyweight: requiring either changing the language runtime or the
application code (e.g., by using monads). In this paper, we present
\racets: an extension of Racket to faceted execution a library of
macros. Given Racket's highly-expressive macro system, our
implementation follows relatively directly from the semantics of
faceted execution. To demonstrate how \racets can be used for
policy-agnostic programming, we use it to build a web-based game of
Battleship. Our implementation sheds light on several interesting
issues in interacting with code written without faceted execution. Our
\racets implementation is open source, under development, and
available online.
\end{abstract}

@@

Policy-agnostic programming promotes the idea that programmers should
be able to write programs ``normally'', without concerning themselves
with how privacy policies are enforced. Unfortunately, the
relatively foundational linguistic changes required to enable faceted
execution have hindered implementations thusfar. Dynamic generation
of first-class security labels, tracking an arbitrary number of
facets per value, and keeping facted-value trees in a canonical order,
are all central challenges in any practical implementation. For example, the
first implementation of FE (by \citet{Austin:2012}) extended
a JavaScript interpreter to account for faceted values. Other
implementations use monads~\cite{Schmitz:2016}
or rely upon third-party macro systems~\cite{Yang:2016}. We
know of no existing implementation of FE that aims to interoperate seamlessly
with code written in the host language. By contrast, Scheme boasts a powerful
hygenic macro system that allows essentially any linguistic form
to be modified arbitrarily.

In this paper we present \racets, an implementation of policy-agnostic
programming in Racket via macros. \racets provides facilities for
creating policies and faceting secure data with those policies. \racets
also extends several core forms in Racket to work with faceted
values (our implementation is detailed in Section~\ref{sec:implementation}). We
have measured the overhead of \racets in several microbenchmarks, and
have used \racets to implement a small server-based
board-game (detailed in Sections~\ref{sec:overview} and~\ref{sec:implementation}). We see \racets as a
promising prototype for policy-agnostic programming in Racket, and
conclude with discussion of future directions in Section~\ref{sec:conclusion}.

@@

To introduce faceted execution more concretely, we present the implementation
of Battleship, a small guessing game, in \racets (this section presents a
distilled version of our case study in
Section~\ref{sec:implementation}). In this game each player has a
private board of coordinates, at which they place ``ships''. The players
hide their boards from each other as play progresses in rounds. Each
turn a player guesses the position of a ship on the other player's
board. If the guess is successful the tile is removed from the
board and a hit is declared publically. Play ends once one player's board
has no remaining tiles, at which point that player loses.

@@

In the above example, \blue{$\star$} (lazy failure) is used in the negative branch
to ensure execution will fail if Bob tries to observe Alice's secret gameboard.
The $\code{(obs}\ \blue{e_\ell}\ \blue{e_\text{key}}\ \blue{e_\text{fac}}\code{)}$ form takes
a label, a key, and a facted value. Explicit observation projects a single label
$\blue{e_\ell}$ in faceted value $\blue{e_\text{fac}}$ to either its positive or negative facet,
depending on whether the policy associated with $\blue{e_\ell}$ returns true for key
$\blue{e_\text{key}}$. If Bob tries to observe Alice's board, the policy predicate
will return false (from \texttt{(= 1 $\blue{2}$)}) and Alice's negative facet $\blue{\star}$ will result.

@@

Because the applied function can be stateful, faceted execution also
vmaintains the current privilege level in a \emph{program counter} (PC). The
program counter is a property of the current evaluation context and is used
to build facets when writes are made to the store in a privileged context.
For example, if a stateful function ``splits'' when applied on both the positive and negative
facets of a value faceted by a label $\blue{\ell}$, and on the positive branch
the function uses $\texttt{set!}$ to mutate a variable $x$ from $2$ to $3$,
FE semantics will set $x$ to $\facet{\ell}{3}{2}$ so that the value $3$
cannot be leaked from the secure context (speculative execution under the
$\blue{+\ell}$ facet). This is because, for the duration of the the app ``split'',
the evaluation context records that all values are implicitly guarded by $\blue{+\ell}$
and then $\blue{-\ell}$, respectively. If the semantics for $\texttt{set!}$ does
not make this faceting explicit, a senstive value could leak from one PC to another.
We expand upon these subtleties in Section~\ref{sec:semantics},
where we present a complete semantics for faceted execution.

@@

Our semantics is shown in Figure~\ref{fig:semantics} \cite{Micinski:2019}.
As \lang is an
extension of the lambda calculus with references, we present the parts
unique to faceted execution in \fcol{red}, while keeping the lambda
calculus with references in \blue{blue}. Base values in our semantics
include addresses (used for boxes), constants, and closures. We also
include a kind of lazy failure (\bmth{\star}), which is necessary for
defining store update within a protected context.

@@

The rules \textsc{Box}, \textsc{Unbox}, and \textsc{Set} appear
similar to the standard implementation of boxes, but employ several
meta-functions to do their work. This is because box creation, reads,
and writes may occur within a privilaged context, and care must be
taken to form facets when \rmth{pc} is nonempty. To understand why,
consider the following example\footnote{Our formal semantics elides
  \code{if}, though it may be obtained via a Church encoding if
  desired as in Austin et al.\cite{Austin:2012}. Our implementation
  includes \code{if} but not other constructs such as \code{cond}}:

@@

Facet creation appears as three separate rules in
Figure~\ref{fig:semantics}, which we recapitulate here in three
distint colors for each case:

@@

Our hypothesis is that this tagging scheme is key to permitting inter-operation
between \racets and Racket, and we have implemented a faceted, web-based game of Battleship
to explore this idea. We suspect that a more thorough investigation of likely idioms
for faceted, non-faceted module interaction is needed and that purely functional code
plays a special role as a degenerate case where arbitrary non-faceted code may be lifted
to operate over facted values without potential unsoundness. In the future, we plan to explore
these design choices in a more principled way and also to scale a static analysis \cite{Micinski:2019}
of faceted execution to fully expanded Racket so it may be applied directly to \racets.
